Toward nonlinear multigrid \\ for nonlinear and non-local variational inequalities

### 2023-02-08 talk/slides.tex
more talk work ... more work todo

\section{variational inequalities (VIs)?}

@@

\section{full approximation scheme (FAS) multigrid for PDEs?}

@@

\begin{itemize}
\item let's not get stuck on textbook example problems!
\item multigrid for a real-world VI problem?
\item consider the steady state of a viscous layer of surface elevation $s(x,y)$ flowing with velocity $\bu(x,y,z)$ under its own weight, over fixed bed topography with elevation $b(x,y)$, in a \emph{climate} which adds or removes fluid at a signed rate $a(x,y)$ [$\text{m}\,\text{s}^{-1}$]
    \begin{itemize}
    \item[$\circ$] data $a,b$ defined on domain $\Omega \subset \RR^2$ (left figure)
    \end{itemize}
\item geophysical examples: \alert{glaciers and ice sheets}, sea ice, oceans and lakes
\end{itemize}

@@

%\vspace{-5mm}
{\footnotesize
\begin{itemize}
\item<2> more ice sheet modeling at my Math.~Geoscience Seminar tomorrow 2pm L5
\end{itemize}
}
\end{frame}

\begin{frame}{strong form}

\begin{itemize}
\item inequality constraint generates free boundary: $s \ge b$
    \begin{itemize}
    \item[$\circ$] ablative climate $a(x) < 0$ forces surface down to bed
    \item[$\circ$] $s\ge b$ $\iff$ thickness $s-b$ must be nonnegative
    \end{itemize}
\item naive strong form of the steady model: % ($\bn_s = \left<-s_x,-s_y,1\right>$ is surface normal):
\begin{align*}
s &\ge b                    & &\text{everywhere in } \Omega \\
-\bu|_s \cdot \bn_s &= a    & &\text{where } s(x) > b(x)
\end{align*}

    \begin{itemize}
    \item[$\circ$] surface velocity $\bu|_s$ is determined by fluid domain geometry
    \item[$\circ$] true $\Phi(s) = -\bu|_s \cdot \bn_s$ is a \emph{non-local} function of $s$
    \end{itemize}
\end{itemize}

@@

\begin{frame}{$\Phi(s) = - \bu|_s \cdot \bn_s$ for glacier ice?}

@@

solve the Stokes problem for $\bu,p$, then evaluate at surface:
    $$\int_{\Lambda = \{b < z <s\}} 2 \nu(D\bu) D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0 \quad \forall v,q$$
    $$\Phi(s) = - \bu|_s \cdot \bn_s$$

    \begin{itemize}
    \item[$\circ$] also extend $\Phi(s)$ by 0 to all of $\Omega$
    \item[$\circ$] non-Newtonian viscosity: $\nu(D\bu) = \frac{1}{2} \Gamma |D\bu|^{\pp-2}$ with e.g.~$\pp = \frac{4}{3}$
    \item[$\circ$] well-posed problem for $\bu,p$
    \item[$\circ$] near-optimal solvers available (Isaac et al 2015)
    \end{itemize}
}

\only<2>{
\item \alert{lubrication approximation model}

apply a nonlinear differential operator to $s$:
    $$\Phi(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_{\bx} s|^{\qq} - \grad_{\bx} \cdot\left(\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_{\bx} s|^{\qq-2} \grad_{\bx} s\right)$$
where $\qq = \nn+1$ and $\gamma>0$ is related to ice softness and density
    \begin{itemize}
    \item e.g.~$\qq = 4$
    \item $\Phi(s)$ is a nonlinear differential operator only because membrane (longitudinal) stresses are not balanced in lubrication approximation
    \end{itemize}
}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item admissible surface elevations: \quad $\mathcal{K} = \left\{r \in \mathcal{V} \,:\, r \ge b\right\}$

    \begin{itemize}
    \item[$\circ$] $\mathcal{V}$ to be determined by viscosity model\footnote{in lubrication approximation for glaciers, $u^{8/3} \in W^{1,4}(\Omega)$ (Jouvet \& Bueler, 2012)}
    \end{itemize}
\item VI problem:

@@

\begin{frame}{the Stokes problem for ice}

\begin{itemize}
\item non-shallow model solves a Stokes problem at each step:
\begin{align*}
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s$}
\end{align*}
\item this is the \alert{stress balance} (conservation of momentum) problem which determines velocity $\bu$ and pressure $p$
\end{itemize}
\end{frame}

\section{scalable approaches for VIs}

@@

\begin{itemize}
\item multigrid for the classical obstacle problem?
    \begin{itemize}
    \item[$\circ$] a nonlinear problem, though $-\nabla^2$ is linear
    \end{itemize}
\item Newton-multigrid straightforward in PETSc:
\begin{Verbatim}[xleftmargin=13mm,fontsize=\scriptsize]
./obstacle -da_grid_x 3 -da_grid_y 3 \
    -snes_type vinewtonrsls -ksp_type cg \
    -pc_type mg -da_refine J
\end{Verbatim}
    \begin{itemize}
    \item[$\circ$] linear solver applies to inactive variables
    \end{itemize}
\item the outer Newton iteration must determine the active set \alert{before} V-cycle multigrid can provide effective preconditioning
    \begin{itemize}
    \item[$\circ$] growing Newton (SNES) iterations
    \end{itemize}
\end{itemize}

@@

\section{scalable approaches for \emph{nonlocal} VIs}

@@

\begin{frame}{\emph{background} references}

>>>>

\section{variational inequalities (VIs)}

@@

\section{full approximation scheme (FAS) multigrid for PDEs}

@@

\begin{itemize}
\item let's not get stuck on textbook example problems!
\item multigrid for a real-world VI problem?
\item consider a viscous layer with surface elevation $s(x,y)$ flowing with velocity $\bu(x,y,z)$, driven by gravity, over fixed bed topography with elevation $b(x,y)$, in a \emph{climate} which adds or removes fluid at a signed rate $a(x,y)$ [$\text{m}\,\text{s}^{-1}$]
    \begin{itemize}
    \item[$\circ$] data $a,b$ defined on domain $\Omega \subset \RR^2$
    \end{itemize}
\item geophysical examples: \alert{glaciers and ice sheets}, sea ice, lakes
\end{itemize}

@@

%\vspace{-5mm}
{\footnotesize
\begin{itemize}
\item<2> more ice sheet modeling at my Math.~Geosci.~Seminar tomorrow 2pm L5
\end{itemize}
}
\end{frame}

\begin{frame}{naive strong form}

\begin{itemize}
\item naive strong form of the steady model: % ($\bn_s = \left<-s_x,-s_y,1\right>$ is surface normal):
\begin{align*}
s &\ge b                    & &\text{everywhere in } \Omega \\
-\bu|_s \cdot \bn_s &= a    & &\text{where } s(x,y) > b(x,y)
\end{align*}

    \begin{itemize}
    \item[$\circ$] surface velocity $\bu|_s$ is determined by fluid domain geometry, including $s$ itself
    \item[$\circ$] $\bn_s=\left<-\grad s,1\right>$ is upward surface normal
    \item[$\circ$] generally: $-\bu|_s \cdot \bn_s$ is a \emph{non-local} function of $s$
    \end{itemize}
\item<2> the inequality constraint $s \ge b$ \alert{generates a free boundary}

where the ablative climate $a < 0$ forces surface down to bed
\end{itemize}

@@

\begin{frame}{how to evaluate $\Phi(s) = - \bu|_s \cdot \bn_s$ for glacier ice?}

@@

solve the Stokes problem, then evaluate velocity at surface:
    $$\int_{\Lambda(s) = \{b < z <s\}} 2 \nu(D\bu) D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv = 0 \quad \forall \bv,q$$
    $$\Phi(s) = - \bu|_s \cdot \bn_s$$

    \begin{itemize}
    \item[$\circ$] assuming incompressibility and non-Newtonian viscosity: $\nu(D\bu) = \frac{1}{2} \Gamma |D\bu|^{\pp-2}$ with $\pp = \frac{4}{3}$
    \item[$\circ$] given $s$, this is a well-posed problem for velocity $\bu\in \mathbf{W}^{1,\pp}$ and pressure $p \in L^\qq$ on domain $\Lambda(s)$
    \item[$\circ$] near-optimal solvers available (Isaac et al 2015)
    \end{itemize}
}

\only<2>{
\item \alert{lubrication approximation}\footnote{also known as the \emph{shallow ice approximation}} \alert{model}

apply a nonlinear elliptic differential operator to $s$:
    $$\Phi(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad s|^{\qq} - \grad \cdot\left(\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad s|^{\qq-2} \grad s\right)$$
    \begin{itemize}
    \item $\qq = 4$
    \item $\grad$ is in $x,y$ only
    \item $\Phi(s)$ is a nonlinear differential operator because membrane stresses are \emph{not} balanced in the approximation
    \item $\Phi(s)$ is doubly-degenerate
    \end{itemize}
}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item admissible surface elevations:
    $$\mathcal{K} = \left\{r \in \mathcal{V} \,:\, r \ge b\right\}$$

    \begin{itemize}
    \item[$\circ$] $\mathcal{V}$ to be determined by viscous fluid model\footnote{in shallow ice approximation, $(s-b)^{8/3} \in W^{1,4}(\Omega)$ (Jouvet \& Bueler, 2012)}
    \end{itemize}
\item VI problem for surface elevation $s\in\mathcal{K}$:
	$$\ip{\Phi(s)}{r-s} \ge \ip{a}{r-s} \quad \text{ for all } r \in \mathcal{K}$$
where
    $$\Phi(s)=- \bu|_s \cdot \bn_s,$$
with extension by 0 to all of $\Omega$, and $\bu$ is the velocity solution on
    $$\Lambda(s) = \{(x,y,z) : b(x,y) < z < s(x,y)\}$$

@@

\begin{frame}{CP form (glacier ice in a climate)}

\begin{itemize}
\item VI form on previous slide is too abstracted for clarity
\item the strong form of the same problem is a \alert{complementarity problem (CP) coupled to a Stokes problem}:
\begin{align*}
s - b &\ge 0 && \text{ in $\Omega$} \\
- \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
(s-b)(- \bu|_s \cdot \bn_s - a) &= 0 && \text{''} \\
- \nabla \cdot \left(2 \nu(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(s)$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\left(2 \nu(D\bu) D\bu - pI\right) \bn &= \bzero && \{z=s\} \subset \partial \Lambda(s) \\
\bu &= \bzero && \{z=b\} \subset \partial \Lambda(s)
\end{align*}
\item solve this for $s$ on $\Omega$, and simultaneously for $\bu,p$ on $\Lambda(s) = \{(x,y,z) : b(x,y) < z < s(x,y)\}$
\end{itemize}
\end{frame}

\begin{frame}{a non-local VI problem}

\begin{itemize}
\item in the Stokes case, the residual $r(s) = a - \Phi(s)$ $= a + \bu|_s \cdot \bn_s$ \alert{depends non-locally on $s$}
\item for example, consider $\bu_{(s+\psi)} - \bu_{(s)}$ from surface perturbation

(hat function) $\psi$
\end{itemize}

\medskip
\begin{center}
\includegraphics[width=0.85\textwidth]{images/stokes-greens-arndt.png}
\end{center}

\vspace{-5mm}
\hfill {\tiny P.~Arndt figure}
\end{frame}

\section{multigrid approaches for VIs?}

@@

\begin{itemize}
\item multigrid for the classical obstacle problem?
    \begin{itemize}
    \item[$\circ$] a nonlinear problem, though $-\nabla^2$ is linear
    \end{itemize}
\item Newton-multigrid is straightforward in PETSc:
\begin{Verbatim}[xleftmargin=13mm,fontsize=\scriptsize]
./obstacle -da_grid_x 3 -da_grid_y 3 \
    -snes_type vinewtonrsls -ksp_type cg \
    -pc_type mg -da_refine J
\end{Verbatim}
    \begin{itemize}
    \item[$\circ$] CG-GMG V-cycles linear solver applies to inactive variables
    \end{itemize}
\item the outer Newton iteration must converge on the active set \alert{before} multigrid can provide effective preconditioning
    \begin{itemize}
    \item[$\circ$] thus growing Newton (SNES) iterations!
    \end{itemize}
\end{itemize}

@@

\section{multigrid approaches for \emph{nonlocal} VIs?}

@@

\begin{frame}{additional background references}
